---
title: "Chapter 5 exercises: Data transformation"
output: 
  html_notebook:
    toc: TRUE
---
```{r}
library(nycflights13)
library(tidyverse)
library(dplyr)
```

## 5.2 Filter rows with filter()

### 5.2.1
> Find all flights that:

> 1. Had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 2)
```
> 2. Flew to Houston (IAH or HOU)

```{r, eval=FALSE}
filter(flights, dest == "IAH" | dest == "HOU")
```  
or  

```{r, eval= TRUE}
filter(flights, dest %in% c("IAH", "HOU"))
```
> 3. Were operated by United, American, or Delta  

```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```
> 4. Departed in summer (July, August, and September)

```{r}
filter(flights, month %in% c(7, 8, 9))
```
> 5. Arrived more than two hours late, but didn’t leave late  

```{r}
filter(flights, dep_delay <= 0 & arr_time >= 120)
```
> 6. Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
filter(flights, dep_delay >= 60 & dep_delay - arr_delay >30 )
```
> 7. Departed between midnight and 6am (inclusive)  

```{r}
summary(flights$dep_time)
```


```{r, eval=TRUE}
filter(flights, dep_time %% 2400 <= 600)
```
### 5.2.2
> Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

This is a shortcut for x >= left & x <= right  
`between(vector of values, left boundary, right boundary)`

```{r, echo=TRUE}
filter(flights, between(month, 7, 9))
```
### 5.2.3

> How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
count(flights, is.na(dep_time))
```

```{r}
summary(flights)
```
## 5.3 Arrange rows with arrange()

### 5.3.1
> How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na())`.  

```{r}
arrange(flights, desc(is.na(dep_time)))
```
### 5.3.2 
> Sort `flights` to find the most delayed flights. Find the flights that left earliest

```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```
### 5.3.3
> Sort flights to find the fastest (highest speed) flights.

```{r}
arrange(flights, desc(distance / ((hour * 60) + minute)))
```
### 5.3.4
> Which flights travelled the farthest? Which travelled the shortest?

```{r}
arrange(flights, desc(distance))
arrange(flights, distance)
```

## 5.4 Select colums with select()

### 5.4.1

>Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, starts_with('dep'), starts_with('arr'))
select(flights, 4, 6, 7, 9)

```
### 5.4.2
>What happens if you include the name of a variable multiple times in a select() call?

```{r}
select(flights, dep_time, dep_time)
```
### 5.4.3

>What does the any_of() function do? Why might it be helpful in conjunction with this vector?

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
flights %>% select(any_of(vars))
```
### 5.4.4

> Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

Default is to ignore case. 

```{r}
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = FALSE))
```
## 5.5 Add new variables with `mutate()`

### 5.5.1

> Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r}
transmute(flights,
          dep_time = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
          sched_dep_time = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1400)
```
### 5.5.2

> Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
select(mutate(flights, arr_dep_time = arr_time - dep_time), air_time, arr_dep_time)

mutate(flights, 
       arr_time = (arr_time %/% 100 * 60 + arr_time %% 100) %% 1440,
       dep_time = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440) %>% 
  transmute(air_time, arr_dep_time = arr_time - dep_time)
```
Still doesn't solve, other variables could be time zone differences.

### 5.5.3

> Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

```{r}
select(flights, dep_time, sched_dep_time, dep_delay)
```
`dep_time - sched_dep_time == dep_delay`

### 5.5.4

> Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().

`min_rank()` assigns tied values the same rank. 
```{r}
min_rank(c(10, 5, 1, 5, 5)) 
```

```{r}
mutate(flights, dep_delay_min_rank = min_rank(desc(dep_delay))) %>% 
  arrange(dep_delay_min_rank)
```
### 5.5.5

> What does `1:3 + 1:10` return? Why?

```{r, warning=TRUE}
1:3 + 1:10
```
You can only add vectors of different lengths if one is a multiple of another.
```{r}
1:2 + 1:10
```
### 5.5.6

> What trigonometric functions does R provide?

These can be viewed in `?Trig` documentation.  
- `cos(x)`, `sin(x)`, `tan(x)`
- `acos(x)`, `asin(x)`, `atan(x)`, `atan2(y, x)` 
- `cospi(x)`, `sinpi(x)`, `tanpi(x)`

## 5.6 Grouped summaries with `summarise()`


### 5.6.1
> Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:
1. A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
2. A flight is always 10 minutes late.
3. A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
4. 99% of the time a flight is on time. 1% of the time it’s 2 hours late.

### 5.6.2

> Come up with another approach that will give you the same output `as not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` (without using `count()`).

`df count(x, wt)` is the same as
`df %>% group_by(x) %>% summarise(n = sum(wt))`  
Adds the total tally of distance for each tailnum in this example

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(arr_delay), !is.na(dep_delay)) #filter out missing number

not_cancelled %>% count(dest)
not_cancelled %>% count(tailnum, wt = distance)
```


```{r}
not_cancelled <- flights %>% 
  filter(!is.na(arr_delay), !is.na(dep_delay)) #filter out missing numbers
  
not_cancelled %>%   
  group_by(dest) %>% #group by destination
  summarise(n()) # summarise counts by destination

not_cancelled %>%   
  group_by(tailnum) %>% #group by destination
  summarise(n = sum(distance))

```
### 5.6.3

> Our definition of cancelled flights is slightly suboptimal. Why? Which is the most important column? 

```{r}
filter(flights, !is.na(dep_delay), is.na(arr_delay)) %>%
  select(dep_time, arr_time, sched_arr_time, dep_delay, arr_delay)
```
### 5.6.4

> Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
(flights %>% 
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>% 
    group_by(year, month, day) %>% 
    summarise(
      cancelled_num = sum(cancelled),
      flights_num = n(),
      cancelled_prop = mean(cancelled)
    )
)
```
```{r}
cancelled_and_delays <- 
  flights %>% 
  mutate(cancelled = is.na(arr_delay) | is.na(dep_delay)) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_prop = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(cancelled_and_delays) + 
  geom_point(aes(cancelled_prop, avg_arr_delay)) +
  coord_flip()
```
### 5.6.5

> Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`)

Florida airlines

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  arrange(desc(avg_arr_delay))
```
### 5.6.6
 > What does the sort argument to count() do? When might you use it?
 
Sorts `count()` in order of `n`. Basically shorthand of `count() %>% arrange()`

## 5.7 Grouped mutates (and filters)

### 5.7.1

### 5.7.2

> Which plane (`tailnum`) has the worst on-time record ? 

```{r}
flights %>% 
  group_by(tailnum) %>% 
  arrange(desc(arr_delay)) %>% 
  select(tailnum, arr_delay)
```













