---
title: 'Chapter 3: Data Visualisation'
output:
  html_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: sentence
---

```{r include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
```

# 3. Data Visualisation

------------------------------------------------------------------------

## 3.2 First Steps

### 3.2.1

> Run `ggplot(data = mpg)`.
> What do you see?

```{r}
ggplot(data = mpg)
```

### 3.2.2

> How many rows are in mpg?
> How many columns?

```{r}
dim(mpg)
```

```{r}
head(mpg)
```

234 observations (rows), 11 variables (columns)

### 3.2.3

> What does the drv variable describe?
> Read the help for `?mpg` to find out.

`?mpg`

drv = the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd.

### 3.2.4

> Make a scatterplot of hwy vs cyl.

```{r}
mpg %>% 
  ggplot(aes(x = hwy, y = cyl)) + 
  geom_point()
```

### 3.2.5

> What happens if you make a scatterplot of class vs drv?
> Why is the plot not useful?

```{r}
mpg %>% 
  ggplot(aes(x = class, y = drv)) + 
  geom_point()
```

------------------------------------------------------------------------

## 3.3 Mapping Aesthetics

### 3.3.1

> What's gone wrong with this code?
> Why are the points not blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

Colour is within the aesthetic mapping of the data.
It should be this:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), colour = "blue")
```

### 3.3.2

> Which variables in mpg are categorical?
> Which variables are continuous?
> (Hint: type `?mpg` to read the documentation for the dataset).
> How can you see this information when you run mpg?

```{r}
mpg
```

displ (engine displacement), cty (city miles per gallon), and hwy (highway miles per gallon) are all continuous variables.
The rest are categorical.

### 3.3.3

> Map a continuous variable to color, size, and shape.
> How do these aesthetics behave differently for categorical vs. continuous variables?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = cty))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cty))
```

```{r, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = cty))
```

### 3.3.4

> What happens if you map the same variable to multiple aesthetics?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = hwy, colour = displ))
```

This mapping is redundant.
One variable should be assigned to one mapping.

### 3.3.5

> What does the stroke aesthetic do?
> What shapes does it work with?
> (Hint: use `?geom_point`)

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 21, colour = "purple", fill = "green", size = 5, stroke = 5)
```

Changes border of points.
geom_point also has other aesthetics such as colour, shape, fill, etc.

### 3.3.6

> What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ \< 5)`?
> Note, you'll also need to specify x and y.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ > 5))
```

Colours observations according to boolean values (TRUE vs FALSE)

------------------------------------------------------------------------

## 3.5 Facets

### 3.5.1

> What happens if you facet on a continuous variable?

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cty)
```

The contiuous variable is converted into a categorical variable and a facet is plotted for each distinct value.

### 3.5.2

> What do the empty cells in plot with `facet_grid(drv \~ cyl)` mean?
> How do they relate to this plot?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty)) +
  facet_grid(drv ~ cyl)
```

The empty facets correlate to missing points in the first plot.

### 3.3.3

> What plots does the following code make?
> What does .
> do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

. ignores that dimension when facetting.
The first `facet_grid(hwy \~ .)` creates facets along the y-axis, the second `facet_grid(. \~ cyl)` creates facets along the x-axis.

### 3.3.4

> Take the first faceted plot in this section:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

> What are the advantages to using faceting instead of the colour aesthetic?
> What are the disadvantages?
> How might the balance change if you had a larger dataset?

Colour aesthetics make it easier to view differences across x and y axis.
Whereas faceting makes it easier to look at a higher number of categories since each one is mapped on a different plot.

### 3.5.5

> Read `?facet_wrap`.
> What does nrow do?
> What does ncol do?
> What other options control the layout of the individual panels?
> Why doesn't `facet_grid()` have nrow and ncol arguments?

nrow and ncol respectively sets number of rows and columns to facet wrap.
facet_grid doesn't have this as these are taken from the arguments for x and y axis.

### 3.5.6

> When using `facet_grid()` you should usually put the variable with more unique levels in the columns.
> Why?

More space for columns in landscape images.

------------------------------------------------------------------------

## 3.6 Geometric Objects

### 3.6.1

> What geom would you use to draw a line chart?
> A boxplot?
> A histogram?
> An area chart?

-   line chart: `geom_line()`
-   boxplot: `geom_boxplot()`
-   histogram: `geom_histogram()`
-   area chart: `geom_area()`

### 3.6.2

> Run this code in your head and predict what the output will look like.
> Then, run the code in R and check your predictions.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

### 3.6.3

> What does `show.legend = FALSE` do?
> What happens if you remove it?
> Why do you think I used it earlier in the chapter?

Removes the legend box.

### 3.6.4

> What does the se argument to `geom_smooth()` do?

Shows the standard error bands.

### 3.6.5

> Will these two graphs look different?
> Why/why not?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

geom_point and geom_smooth inherits the same mappings from ggplot(mapping = aes())

### 3.6.6

> Recreate the R code necessary to generate the following graphs.

```{r message=FALSE, warning=FALSE}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy, line = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy, colour = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy, colour = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy, colour = drv, colour = drv, linetype = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy, colour = drv)) + 
  geom_point(size = 4, colour = "white") +
  geom_point(aes(colour = drv))
  
```

------------------------------------------------------------------------

## 3.7 Statistical Transformations

### 3.7.1

> What is the default geom associated with `stat_summary()`?
> How could you rewrite the following plot to use that geom function instead of the stat function?

```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

Default geom for `stat_summary()` is `geom_pointrange`, but the default stat for `geom_pointrange` is 'identity'.

```{r}
ggplot(data = diamonds) + 
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = 'summary',
    fun.max = max,
    fun.min = min,
    fun = median
  )
   
```

### 3.7.2

> What does `geom_col()` do?
> How is it different to `geom_bar()`?
> `geom_col()` by default uses `stat_identity()`, leaves the data as is and expects both `x` and `y` mappings.
> `geom_bar` by default uses `stat_count`, only expects an `x` mapping and calculates the number of observations for each value of x.

### 3.7.3

> Most geoms and stats come in pairs that are almost always used in concert.
> Read through the documentation and make a list of all the pairs.
> What do they have in common?

| geom                    | stat                    |
|-------------------------|-------------------------|
| `geom_bar()`            | `stat_count()`          |
| `geom_bin2d()`          | `stat_bin_2d()`         |
| `geom_boxplot()`        | `stat_boxplot()`        |
| `geom_contour_filled()` | `stat_contour_filled()` |
| `geom_contour()`        | `stat_contour()`        |
| `geom_count()`          | `stat_sum()`            |
| `geom_density_2d()`     | `stat_density_2d()`     |
| `geom_density()`        | `stat_density()`        |
| `geom_dotplot()`        | `stat_bindot()`         |
| `geom_function()`       | `stat_function()`       |
| `geom_sf()`             | `stat_sf()`             |
| `geom_sf()`             | `stat_sf()`             |
| `geom_smooth()`         | `stat_smooth()`         |
| `geom_violin()`         | `stat_ydensity()`       |
| `geom_hex()`            | `stat_bin_hex()`        |
| `geom_qq_line()`        | `stat_qq_line()`        |
| `geom_qq()`             | `stat_qq()`             |
| `geom_quantile()`       | `stat_quantile()`       |

: Complementary geoms and stats.

| geom                       | default stat               | shared docs |
|:---------------------------|:---------------------------|-------------|
| `geom_abline()`            | `stat_identity()`          |             |
| `geom_area()`              | `stat_identity()`          |             |
| `geom_bar()`               | `stat_count()`             | x           |
| `geom_bin2d()`             | `stat_bin_2d()`            | x           |
| `geom_blank()`             | None                       |             |
| `geom_boxplot()`           | `stat_boxplot()`           | x           |
| `geom_col()`               | `stat_identity()`          |             |
| `geom_count()`             | `stat_sum()`               | x           |
| `geom_countour_filled()`   | `stat_countour_filled()`   | x           |
| `geom_countour()`          | `stat_countour()`          | x           |
| `geom_crossbar()`          | `stat_identity()`          |             |
| `geom_curve()`             | `stat_identity()`          |             |
| `geom_density_2d_filled()` | `stat_density_2d_filled()` | x           |
| `geom_density_2d()`        | `stat_density_2d()`        | x           |
| `geom_density()`           | `stat_density()`           | x           |
| `geom_dotplot()`           | `stat_bindot()`            | x           |
| `geom_errorbar()`          | `stat_identity()`          |             |
| `geom_errorbarh()`         | `stat_identity()`          |             |
| `geom_freqpoly()`          | `stat_bin()`               | x           |
| `geom_function()`          | `stat_function()`          | x           |
| `geom_hex()`               | `stat_bin_hex()`           | x           |
| `geom_histogram()`         | `stat_bin()`               | x           |
| `geom_hline()`             | `stat_identity()`          |             |
| `geom_jitter()`            | `stat_identity()`          |             |
| `geom_label()`             | `stat_identity()`          |             |
| `geom_line()`              | `stat_identity()`          |             |
| `geom_linerange()`         | `stat_identity()`          |             |
| `geom_map()`               | `stat_identity()`          |             |
| `geom_path()`              | `stat_identity()`          |             |
| `geom_point()`             | `stat_identity()`          |             |
| `geom_pointrange()`        | `stat_identity()`          |             |
| `geom_polygon()`           | `stat_identity()`          |             |
| `geom_qq_line()`           | `stat_qq_line()`           | x           |
| `geom_qq()`                | `stat_qq()`                | x           |
| `geom_quantile()`          | `stat_quantile()`          | x           |
| `geom_raster()`            | `stat_identity()`          |             |
| `geom_rect()`              | `stat_identity()`          |             |
| `geom_ribbon()`            | `stat_identity()`          |             |
| `geom_rug()`               | `stat_identity()`          |             |
| `geom_segment()`           | `stat_identity()`          |             |
| `geom_sf_label()`          | `stat_sf_coordinates()`    | x           |
| `geom_sf_text()`           | `stat_sf_coordinates()`    | x           |
| `geom_sf()`                | `stat_sf()`                | x           |
| `geom_smooth()`            | `stat_smooth()`            | x           |
| `geom_spoke()`             | `stat_identity()`          |             |
| `geom_step()`              | `stat_identity()`          |             |
| `geom_text()`              | `stat_identity()`          |             |
| `geom_tile()`              | `stat_identity()`          |             |
| `geom_violin()`            | `stat_ydensity()`          | x           |
| `geom_vline()`             | `stat_identity()`          |             |

: ggplot2 geom layers and their default stats.

| stat                       | default geom            | shared docs |
|:---------------------------|:------------------------|-------------|
| `stat_bin_2d()`            | `geom_tile()`           |             |
| `stat_bin_hex()`           | `geom_hex()`            | x           |
| `stat_bin()`               | `geom_bar()`            | x           |
| `stat_boxplot()`           | `geom_boxplot()`        | x           |
| `stat_count()`             | `geom_bar()`            | x           |
| `stat_countour_filled()`   | `geom_contour_filled()` | x           |
| `stat_countour()`          | `geom_contour()`        | x           |
| `stat_density_2d_filled()` | `geom_density_2d()`     | x           |
| `stat_density_2d()`        | `geom_density_2d()`     | x           |
| `stat_density()`           | `geom_area()`           |             |
| `stat_ecdf()`              | `geom_step()`           |             |
| `stat_ellipse()`           | `geom_path()`           |             |
| `stat_function()`          | `geom_function()`       | x           |
| `stat_function()`          | `geom_path()`           |             |
| `stat_identity()`          | `geom_point()`          |             |
| `stat_qq_line()`           | `geom_path()`           |             |
| `stat_qq()`                | `geom_point()`          |             |
| `stat_quantile()`          | `geom_quantile()`       | x           |
| `stat_sf_coordinates()`    | `geom_point()`          |             |
| `stat_sf()`                | `geom_rect()`           |             |
| `stat_smooth()`            | `geom_smooth()`         | x           |
| `stat_sum()`               | `geom_point()`          |             |
| `stat_summary_2d()`        | `geom_tile()`           |             |
| `stat_summary_bin()`       | `geom_pointrange()`     |             |
| `stat_summary_hex()`       | `geom_hex()`            |             |
| `stat_summary()`           | `geom_pointrange()`     |             |
| `stat_unique()`            | `geom_point()`          |             |
| `stat_ydensity()`          | `geom_violin()`         | x           |

: ggplot2 stat layers and their default geoms.

### 3.7.4

> What variables does `stat_smooth()` compute?
> What parameters control its behaviour?

```{r, eval=FALSE}
?stat_smooth
```

`geom_smooth` and `stat_smooth` are aliases and use the same arguments.
These methods/arguments are: - `method`: smoothing method.
e.g.
`NULL`, `lm`, `glm`, `loess`.
- `formula`: customiing smoothing method - `se`: If `TRUE`, displays standard error bands.
- `na.rm`: missing values are removed with `FALSE` or without `TRUE` error messages.

Computed variables of `stat_smooth` include: - `y`: predicted values - `ymin`: lower value of the confidence interval - `ymax`: upper value of confidence interval - `se`: standard error

### 3.7.5

> In our proportion bar chart, we need to set group = 1.
> Why?
> In other words what is the problem with these two graphs?

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```

From `?geom_bar`: `geom_bar()` makes the height of the bar proportional to the number of cases in each group.
Proportions are calculated within each group so they all have the same height.
The correct plot is:

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

------------------------------------------------------------------------

## 3.8 Position adjustments

### 3.8.1

> What is the problem with this plot?
> How could you improve it?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

There are missing observations in this plot because they are stacked.
We can apply a jitter (random noise displacement) to each observation so they are less likely to stack.
This reduces accuracy.
Use `geom_jitter` or add a `position = "jitter"` argument to `geom_point`.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter" )
```

### 3.8.2

> What parameters to `geom_jitter()` control the amount of jittering?

-   `width`: controls horizontal jitter
-   `height`: controls vertical jitter

### 3.8.3

> Compare and contrast `geom_jitter()` with `geom_count`.

-   `geom_jitter` adds random noise to displace observations on the graph, thereby reducing overplotting but also accuracy.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter()
```

-   `geom_count` sizes points according to relative number of observations. eg.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```

### 3.8.4

> What's the default position adjustment for geom_boxplot()?
> Create a visualisation of the mpg dataset that demonstrates it.

The default position of `geom_boxplot` is `"dodge2"`.
This parameter allows moves geoms so that they do not overlap.

```{r}
ggplot(data = mpg, aes(x = drv, y = hwy, colour = class)) +
  geom_boxplot()

ggplot(data = mpg, aes(x = drv, y = hwy, colour = class)) +
  geom_boxplot(position = "identity")
```

------------------------------------------------------------------------

## 3.9 Coordinate systems

### 3.9.1

> Turn a stacked bar chart into a pie chart using `coord_polar()`.

```{r}
ggplot(diamonds, aes(x = factor(1), fill = cut)) + 
  geom_bar()

ggplot(diamonds, aes(x = factor(1), fill = cut)) + 
  geom_bar() + 
  coord_polar(theta = "y")
```

### 3.8.2

> What does `labs()` do?
> Read the doc.

Modifies axis, legend and plot labels.
- `title`: The text for the title.
- `subtitle`: The text for the subtitle for the plot which will be displayed below the title.
- `caption`:The text for the caption which will be displayed in the bottom-right of the plot by default.
- `tag`: The text for the tag label which will be displayed at the top-left of the plot by default.
- `alt`, `alt_insight`: Text used for the generation of alt-text for the plot.
See get_alt_text for examples.
- `label`: The title of the respective axis (for `xlab()` or `ylab()`) or of the plot (for `ggtitle()`).

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "Highway MPG",
       x = "Class",
       title = "Highway MPG by car class",
       subtitle = "1999-2008",
       caption = "Source: http://fueleconomy.gov")
```

### 3.8.3

> What's the difference between coord_quickmap() and coord_map()?

Both functions project 3D Earth onto a 2D plane.
`coord_map()` uses the Mercator project whereas `coord_quickmap()` uses a faster projection.

### 3.8.4 
> What does the plot below tell you about the relationship between city and highway mpg? Why is coord_fixed() important? What does geom_abline() do?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

cty and hwy are linearly dependent.
`coord_fixed()` fixes the aspect ratio so `geom_abline()` is displayed at 45 degrees.
