---
title: "Chapter 5 exercises: Data transformation"
output: 
  html_notebook:
    toc: TRUE
---
```{r}
library(nycflights13)
library(tidyverse)
library(dplyr)
```

## 5.2 Filter rows with filter()

### 5.2.1
> Find all flights that:

> 1. Had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 2)
```
> 2. Flew to Houston (IAH or HOU)

```{r, eval=FALSE}
filter(flights, dest == "IAH" | dest == "HOU")
```  
or  

```{r, eval= TRUE}
filter(flights, dest %in% c("IAH", "HOU"))
```
> 3. Were operated by United, American, or Delta  

```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```
> 4. Departed in summer (July, August, and September)

```{r}
filter(flights, month %in% c(7, 8, 9))
```
> 5. Arrived more than two hours late, but didn’t leave late  

```{r}
filter(flights, dep_delay <= 0 & arr_time >= 120)
```
> 6. Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
filter(flights, dep_delay >= 60 & dep_delay - arr_delay >30 )
```
> 7. Departed between midnight and 6am (inclusive)  

```{r}
summary(flights$dep_time)
```


```{r, eval=TRUE}
filter(flights, dep_time %% 2400 <= 600)
```
### 5.2.2
> Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?

This is a shortcut for x >= left & x <= right  
`between(vector of values, left boundary, right boundary)`

```{r, echo=TRUE}
filter(flights, between(month, 7, 9))
```
### 5.2.3

> How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
count(flights, is.na(dep_time))
```

```{r}
summary(flights)
```


